Sexual selection and speciation: a macro-evolutionary approach

### 2018-09-14 Sexual Selection and Extinction.Rmd
Long time no update

---
title: 'Sexual selection and extinction: a macro-evolutionary approach'
author: "Justin G. Cally^1^, Iliana Medina Guzmán^1^ Luke Holman^1^, Devi Stuart-Fox^1^ <br></br> <br></br> ^1^The University of Melbourne"
subtitle: Supplementary Material
bibliography: References/SS_Extinction.bib
csl: References/Proc_b.csl
output:
  html_document:
    toc: true # table of content true
    toc_float: true # make 
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: cosmo
    code_folding: hide # awesome buttons to show/hide the code
  pdf_document: default
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
```

@@

For this project we aimed to compile data on a large set of species in the Passerine bird order that would allow us to account for many evolutionary and environmental predictors of extinction. The following details the source and use of the data.

##Extinction Risk

To obtain extinction risk estimates we used Bayesian Analysis of Macroevolutionary Mixtures (BAMM). We document the control file and process for obtaining extinction rates from BAMM in an accompanying R markdown in the ``BAMM`` folder. Additionally, we compared our BAMM output files (event data) against a pre-existing extinction rate dataset [@Harvey_2017].

Rather than obtaining one estimate of extinction rate, we ran BAMM XXXX times to obtain uncertainty estimates of the extinction risk generated from the phylogeny created by @Jetz_2012.

@Jetz_2012 also provides us with a compiled IUCN threat level listing for all species from 2010 standings.

@@

Briefly, @Dale_2015 obtained male and female plumage score, which can be divided to obtain a sexual dimorphism index (SDi). The plumage score was scored using the _Handbook of the Birds of the World_ @delhoyo2011. By scanning images of males and females in this book across multiple patches @Dale_2015 were able to obtain mean plumage scores from RGB values. Notably concern about the accuracy of these measurements were raised, thus this dataset was previously compared with a more thorough reflectance dataset of Australian birds, the two datasets correlated but was expectedly noisy (R2 = 0.67, P < 0.0001). Here we compare the @Dale_2015 data against another reflectance dataset from @Armenta_2008 which has colour discriminability, segment classification and PCA.

Alongside sexual dichromatism data there is also partial data for other metrics analysed in @Dale_2015 such as: Body size, tropical life history, Sexual selection, Cooperative breeding and Migration.

@@

We obtained species range distribution maps from Birdlife International (through IUCN). These species range maps cover ~ 7,000 species of birds. From these ranges we were able to obtain estimates of:

* Range Size
* Average and standard deviation in 19 bioclimatic variables (each range randomly sampled 1000 times)
* Average and standard deviation in 19 bioclimatic variables from the last glaciam maximum (LGM)
* Net primary productivity (NPP) estimates and variability
* Average and standard deviation of population density in the species ranges.

The code for extracting the environmental data from the species is provided in another document XXXX.

#Generating biologically relevent predictors of extinction

@@

Extinction rate ~ Sexual selection
                   + Range size
                   + Short temporal variability of temperature (mean BIOCLIM4)
                   + Short temporal variability of precipitation (mean BIOCLIM15)
                   + Spatial variability of temperature (PCA1)
                   + Spatial variability of precipitation (PCA2)
                   + Long-term variability of temperature (BIOCLIM1)
                   + Long-term variability of precipitation (BIOCLIM12)
                   + Variability of NPP

These predictors can be generated from the compiled datasets seen here:

```{r}
complete.dataframe <- read.csv('data/complete.dataframe.csv')
```

@@

```{r}
restricted.data <- complete.dataframe %>% drop_na(bioclim1) #Drop species without environmental data
PCA.bioclim <- prcomp(restricted.data[c(34,35,36,38:47,49:52)], #Select se.bioclim
                      scale = TRUE, center = TRUE)
PCA.predictions <- predict(PCA.bioclim)
restricted.data <- cbind(restricted.data, PCA.predictions)

#run GAM
PC1.gam <- mgcv::gam(PC1 ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian")

@@

#plot
plot.gam(PC1.gam, residuals = T)
```

We can do the same for PC2 (precipitation)

```{r}
PC2.gam <- mgcv::gam(PC2 ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian")

@@

plot.gam(PC2.gam, residuals = T)
```

##Long term climate variability (LGM anomaly)

To gain estimates for change in climate over the past 22,000 we can use the difference in annual mean temperature (BIO1) and annual precipitation (BIO12) between the LGM (CCSM4 models) and present values. Additionally we can calculate the difference in variability based on the standard deviation.

```{r}
#Change in mean temp
restricted.data$BIO1.LGM.Diff <- restricted.data$bioclim1 - restricted.data$cclgmbi1
#Change in variation temp
restricted.data$BIO1.LGM.var.Diff <- restricted.data$se.bioclim1 - restricted.data$se.cclgmbi1

#Change in mean precipitation
restricted.data$BIO12.LGM.Diff <- restricted.data$bioclim12 - restricted.data$cclgmbi12
#Change in variation precipitation
restricted.data$BIO12.LGM.var.Diff <- restricted.data$se.bioclim12 - restricted.data$se.cclgmbi12

#Inspect these changes 
restricted.data %>% ggplot(aes(x=BIO1.LGM.Diff, y = BIO1.LGM.var.Diff))+
  geom_point()

restricted.data %>% ggplot(aes(x=BIO12.LGM.Diff, y = BIO12.LGM.var.Diff))+
  geom_point()
```

From these plots we can see that on average temperature has increased but the variation in tempere has decresed (narrowed). Alternatively, precipitation sometimes increases and sometimes decreases. When precipitation increases there isw also an increase in variation in that range. However, when annual precipitation decreases there is a decrease in variation.

##Variation in NPP

Similar to the PCA's we expect NPP variation to be correlated with range size. To account for this we can run a GAM and obtain residuals.

```{r}
#run gam
NPP.gam <- mgcv::gam(NPP.se ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian")

#obtain residuals
restricted.data$NPP.res <- residuals.gam(NPP.gam)

#plot
plot(NPP.gam, residuals = T)
```

@@

##Preliminary Analysis of data

Now that we have all our relevent predictors we can combine these with the tip extinction rates from BAMM (code for this is provided in the repository). First, let us inspect the tip rates that @Harvey_2017 obtained in BAMM with our predictors.

```{r}
complete.dataframe %>% ggplot(aes(x=rab.ex.rate.lumped, y=rab.ex.rate.split))+
  geom_point()+
  scale_x_log10()+
  scale_y_log10()
```

The extinction rates of the lumped and split datasets show high similarity, insisting that we could use either.

Let us create a relatively simple model
```{r}
model.1 <- lm(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                data = restricted.data)
summary(model.1)
```

Now we can try a PGLS model 
```{r}
hackett.consensus.tree <- read.nexus('bamm_extinction/data/trees/hackett_dale_sub.consensus.nex')
restricted.data <- restricted.data %>% drop_na(rab.ex.rate.split)
rownames(restricted.data) <- restricted.data$TipLabel
pruned.tree<-drop.tip(hackett.consensus.tree,hackett.consensus.tree$tip.label[-match(restricted.data$TipLabel, hackett.consensus.tree$tip.label)])
pgls.model.1 <- gls(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                correlation = corBrownian(phy = pruned.tree), 
                data = restricted.data, 
                method = "REML")
summary(pgls.model.1)
```

Or perhaps OU model rather than Brownian:

```{r}
pgls.model.3 <- gls(rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                correlation = corMartins(1, phy = pruned.tree), 
                data = restricted.data, 
                method = "REML")
summary(pgls.model.3)
```

Now a Bayesian model that uses a covariance matrix from the phylogeny:

```{r}
pruned.tree$edge.length[pruned.tree$edge.length<0]<-1e-9
pruned.tree$edge.length[pruned.tree$edge.length==0]<-1e-9
pruned.tree <- check_and_fix_ultrametric(pruned.tree)
inv.phylo <- MCMCglmm::inverseA(pruned.tree, nodes = "TIPS", scale = TRUE)
A <- solve(inv.phylo$Ainv)
rownames(A) <- rownames(inv.phylo$Ainv)
```

```{r, eval=FALSE}
model_bayesian <- brm(
  rab.ex.rate.split ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res, data = restricted.data, 
  family = gaussian(), 
  cov_ranef = list(phylo = A),
                seed = 1,
                cores = 4, chains = 4, iter = 4000, #Run 4 chains in parallel for 4000 iterations (2000 are burn in)
                control = list(adapt_delta = 0.99, max_treedepth = 15))
```

```{r}
summary(model_bayesian)
```

Try looking at speciation rate using ES-Sim adapted by @Harvey_essim_2017 and found here https://github.com/mgharvey/ES-sim

```{r}
Essim.frame <- restricted.data$SDi
names(Essim.frame) <- restricted.data$TipLabel

essim.results <- essim(pruned.tree, Essim.frame, nsim = 1000, return.es = TRUE)
```

We can then add the essim.results and run a similar PGLS model
```{r}
restricted.data$ESsim <- essim.results[["es"]]

restricted.data %>% ggplot(aes(x=BIO12.LGM.Diff, y = (ESsim))) + 
  geom_point()+
  geom_smooth(method = "lm")
```

 Once we have our EsSim results we can use them in a linear model to assess the relationship between speciation and our predictors

```{r}
model.2 <- lm(ESsim ~ SDi 
                + range.size.m2 
                + bioclim4
                + bioclim15
                + residuals.PC1
                + residuals.PC2
                + BIO1.LGM.Diff
                + BIO12.LGM.Diff
                + NPP.res,
                data = restricted.data)
summary(model.2)
```

Have a look at how these would look if we obtained model predictions for each data point.

```{r, fig.width=10, fig.height=5}
predict.frame <- cbind(predict(model.2),restricted.data)
predict.frame %>% ggplot(aes(y = `predict(model.2)`, x = SDi)) + 
  geom_point(aes(color = BLFamilyLatin))+
  scale_color_hue(h = c(0,100), l = 50) +
  geom_smooth(color = 'black')+
  theme_minimal()
```

It looks like some of the higher SDi's have slightly higher speciation rates. Perhaps we can look at those obviously dimorphic groups as category of its own.

```{r}
restricted.data %>% ggplot(aes(SDi))+
  geom_histogram(binwidth = 0.015)+
  theme_minimal()
```

>>>>

---
title: 'Sexual selection and speciation: a macro-evolutionary approach'
author: "Justin G. Cally^1^, Iliana Medina Guzmán^1^ Luke Holman^1^, Devi Stuart-Fox^1^ <br></br> <br></br> ^1^The University of Melbourne"
subtitle: Supplementary Material
bibliography: References/SS_Extinction.bib
csl: References/AmNat.csl
output:
  html_document:
    toc: true # table of content true
    toc_float: true # make 
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: cosmo
    code_folding: hide # awesome buttons to show/hide the code
  pdf_document: default
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(grid)
library(gridExtra)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
library(purrr)
library(reshape2)
library(ggExtra)
library(grid)
library(dplyr)
library(BAMMtools)
library(gridExtra)
library(car)
library(coda)
library(MuMIn)
library(parallel)
library(HDInterval)
library(phangorn)
library(kableExtra)
library(RColorBrewer)
library(tibble)
library("EBImage") # for images
library("ggtree")
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
source("functions/diversification_rate_calculation.R")
source("functions/Vdodge_function.R")
source('functions/symmetrise_scale.R')
source('functions/Correlation_matrix.R')
source('functions/gheatmap2.R') #Gets rid of spaces between tiles
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

@@

For this project we aimed to compile data on a large set of species in the Passerine bird order that would allow us to account for many evolutionary and environmental predictors of speciation. The following details the source and use of the data.

##Speciation Rate

To obtain speciation rate estimates we used Bayesian Analysis of Macroevolutionary Mixtures (BAMM). Additionally, we compared our BAMM output files (event data) against a pre-existing extinction rate dataset [@Harvey_2017].

Rather than obtaining one estimate of speciation rate from one tree, we ran BAMM 100 times on 100 trees plus an MCC tree to obtain uncertainty estimates of the speciation rate generated from the phylogeny created by @Jetz_2012. We also estimated extinction rate from BAMM.

@@

Briefly, @Dale_2015 obtained male and female plumage score, which can be divided to obtain a sexual dichromatism index (SDi). The plumage score was scored using the _Handbook of the Birds of the World_ @delhoyo2011. By scanning images of males and females in this book across multiple patches @Dale_2015 were able to obtain mean plumage scores from RGB values. Notably concern about the accuracy of these measurements were raised, thus this dataset was previously compared with a more thorough reflectance dataset of Australian birds, the two datasets correlated but was expectedly noisy. Here we compare the @Dale_2015 data against another reflectance dataset from @Armenta_2008 using colour discriminability as the parameter.

Alongside sexual dichromatism data there is also partial data for other metrics analysed in @Dale_2015 such as: Body size, tropical life history, Sexual selection, Cooperative breeding and Migration. We use a phylogenetic PCA (PC1) that estimates male-bias sexual selection from a species level of sexual size dimorphism, social polygyny and paternal care.

@@

We obtained species range distribution maps from Birdlife International (through IUCN) [@birdlife_2017]. These species range maps cover nearly all species of birds. From these spatial information files we were able to obtain estimates of the following (note that not all the data extracted was subsequently used in the analysis):

* Range Size
* Average and standard deviation in 19 bioclimatic variables (each range randomly sampled 1000 times)
* Average and standard deviation in 19 bioclimatic variables from the last-glacial maximum (LGM) and the last-inter-glacial (LIG)
* Net primary productivity (NPP) estimates and variability
* Average and standard deviation of human population density in the species ranges.

The code for extracting the environmental data from the species is provided in another R-markdown document **Github link**, however no raw data is provided alongside this file due to the large file size of the shapefile and raster information. We have put the extracted environmental variables in a csv filed called ``complete.dataframe.csv`` for convenience. Briefly for each of the ~ 6,000 species, we:

**1)** Randomly sampled the range polygon 1000 times:

```{r, eval = FALSE}
#Increse the iterations (defaukt is 4) so we can obtain complete samples of each range
bird.points <- lapply(bird.ranges, 
                       function(x) {spsample(x, n=1000, type="random", iter = 30)})
```

**2)** Extracted the bioclim or other (e.g. NPP) data from each of the points:
```{r, eval = FALSE}
bird.values <- lapply(bird.points, function(x) {raster::extract(bioclim_data, x)})
bird.values <- as.list(data.frame((bird.values)))
```

**3)** Summarised the extracted data across the 1,000 points into a summary value of interest (means and se) for each species and exported that data.

```{r, eval = FALSE}
#Obtain means
bird.frame <- lapply(bird.values, function(x) {as.data.frame(x)})
bird.summary <- lapply(bird.frame, function(x) {
(as.data.frame(apply(x,2,mean, na.rm =T )))})

#transpose
bird.means <- t(as.data.frame(bird.summary))
bird.means <- (split(bird.means, 1:19))

#Now add column of species name: Same order carried through
bird.means <- cbind.data.frame(shps.jetz, bird.means)

rownames(bird.means) <- NULL
colnames(bird.means) <- c("binomial","bioclim1", "bioclim2", "bioclim3", "bioclim4", "bioclim5", "bioclim6", "bioclim7", "bioclim8", "bioclim9", "bioclim10", "bioclim11", "bioclim12", "bioclim13", "bioclim14", "bioclim15", "bioclim16", "bioclim17", "bioclim18", "bioclim19")

#Write csv
write.csv(bird.means, 'data/bird.means.csv')
```

**4)** This process was repeated when we used gridded data for the LGM, LIG, NPP and human population density, while range size was obtained from the following code:

```{r, eval = FALSE}
bird.range.size <- sapply(bird.ranges,
                       function(x) {(area(x))})
bird.range.size <- sapply(bird.range.size, function(x) {sum(x)})
bird.range.size <- as.data.frame(bird.range.size)
bird.range.size <- cbind.data.frame(shps.jetz, bird.range.size)
rownames(bird.range.size) <- NULL
colnames(bird.range.size) <- c("binomial", "range.size.m2")

#write.csv
write.csv(bird.range.size, 'data/bird.range.size.csv')
```

#Generating biologically relevent predictors

@@

 Extinction/Diversification ~ Sexual selection  
                            + Range size  
                            + Short temporal variability of temperature (mean BIOCLIM4)  
                            + Spatial variability of temperature (PCA1) [residual.PCA1]  
                            + Long-term variability of temperature (LIG)  
                            + NPP

These predictors can be obtained from the compiled datasets seen here:

```{r, warning=FALSE}
plumage.scores <- read.csv('data/plumage_scores.csv')
#Generate sexual dichromatism score: 
plumage.scores$SDi <- abs(plumage.scores$Male_plumage_score - plumage.scores$Female_plumage_score)

#Make DF with enviro variables and plumage scores
complete.dataframe <- read.csv('data/complete.dataframe.csv')
complete.dataframe <- left_join(plumage.scores %>% dplyr::select(binomial, Male_plumage_score, Female_plumage_score, SDi), complete.dataframe, by = "binomial")
```

@@

**Table S1:** Loadings for the first two PCs, when we conduct a PCA on the variation (standard error, where n ~ 1,000) pf each bioclim variable except the standard error of temperature seasonality (se.bioclim4) and the standard error of precipitation seasonality (se.bioclim15), as these are already measures of variability and based on other bioclim variables. We use the standard error as, while we attempted to sample each species range 1,000 times, some points returned NA values and would thus affect the standard deviation, this inconvenience should not affect standard error.
```{r}
restricted.data <- complete.dataframe %>% drop_na(bioclim1) #Drop species without environmental data
PCA.bioclim <- prcomp(restricted.data[c('se.bioclim1', 'se.bioclim2', 'se.bioclim3', 'se.bioclim5', 'se.bioclim6', 'se.bioclim7', 'se.bioclim8', 'se.bioclim9', 'se.bioclim10', 'se.bioclim11', 'se.bioclim12', 'se.bioclim13', 'se.bioclim14', 'se.bioclim16', 'se.bioclim17', 'se.bioclim18', 'se.bioclim19')], #Select se.bioclim
                      scale = TRUE, center = TRUE)
PCA.predictions <- predict(PCA.bioclim)
restricted.data <- cbind(restricted.data, PCA.predictions)

#Create table with PC1 and PC2
as.data.frame(PCA.bioclim$rotation[,1:2]) %>% `colnames<-`(c("PC1", "PC2")) %>% pander()
```

```{r, fig.width=10, fig.height=5}
#run GAM
PC1.gam <- mgcv::gam(PC1*(-1) ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian") #Inverse as the PC1 loads to the negative (COUNTER-NTUITIVE)

@@

#We can do the same for PC2
PC2.gam <- mgcv::gam(PC2 ~ s(log(range.size.m2)), data = restricted.data, family = "gaussian")

@@

#plot
par(mfrow = c(1,2))
plot.gam(PC1.gam, residuals = T, main = 'PC1 residuls (temp)')
plot.gam(PC2.gam, residuals = T, main = 'PC2 residuals (precip)')
```

**Figure S1:** The relationship between spatial variability in the temperature components (PC1) and log-range size is relatively strong. But not as strong for spatial variability in precipitation (PC2). In the analysis we only used the residuals from PC1

##Long term climate variability (LIG anomaly)

To gain estimates for change in climate over the past ~130,000 years we can use the difference in bioclim variables between the LIG and present values. The plots show the two PCs, broadly representing temperature and prescipitation. Data has also been provided for the last-glacial maximum (LGM) but was not used in the analysis.

**Table S2:** Loadings for the first two PCs of each of the PCAs for the difference in bioclimatic variables between today and the LIG. Here, PC1 is more heavily loaded for absoluted temperature difference, while PC2 is more heavily loaded for absolute difference in precipitation.

```{r}
historical.variation.data <- as.data.frame(restricted.data[1])

#FOR LIG

historical.variation.data$bio1.LIG.diff <- abs(restricted.data$bioclim1 - restricted.data$LIG.bi1)
historical.variation.data$bio2.LIG.diff <- abs(restricted.data$bioclim2 - restricted.data$LIG.bi2)
historical.variation.data$bio3.LIG.diff <- abs(restricted.data$bioclim3 - restricted.data$LIG.bi3)
historical.variation.data$bio4.LIG.diff <- abs(restricted.data$bioclim4 - restricted.data$LIG.bi4)
historical.variation.data$bio5.LIG.diff <- abs(restricted.data$bioclim5 - restricted.data$LIG.bi5)
historical.variation.data$bio6.LIG.diff <- abs(restricted.data$bioclim6 - restricted.data$LIG.bi6)
historical.variation.data$bio7.LIG.diff <- abs(restricted.data$bioclim7 - restricted.data$LIG.bi7)
historical.variation.data$bio8.LIG.diff <- abs(restricted.data$bioclim8 - restricted.data$LIG.bi8)
historical.variation.data$bio9.LIG.diff <- abs(restricted.data$bioclim9 - restricted.data$LIG.bi9)
historical.variation.data$bio10.LIG.diff <- abs(restricted.data$bioclim10 - restricted.data$LIG.bi10)
historical.variation.data$bio11.LIG.diff <- abs(restricted.data$bioclim11 - restricted.data$LIG.bi11)
historical.variation.data$bio12.LIG.diff <- abs(restricted.data$bioclim12 - restricted.data$LIG.bi12)
historical.variation.data$bio13.LIG.diff <- abs(restricted.data$bioclim13 - restricted.data$LIG.bi13)
historical.variation.data$bio14.LIG.diff <- abs(restricted.data$bioclim14 - restricted.data$LIG.bi14)
historical.variation.data$bio15.LIG.diff <- abs(restricted.data$bioclim15 - restricted.data$LIG.bi15)
historical.variation.data$bio16.LIG.diff <- abs(restricted.data$bioclim16 - restricted.data$LIG.bi16)
historical.variation.data$bio17.LIG.diff <- abs(restricted.data$bioclim17 - restricted.data$LIG.bi17)
historical.variation.data$bio18.LIG.diff <- abs(restricted.data$bioclim18 - restricted.data$LIG.bi18)
historical.variation.data$bio19.LIG.diff <- abs(restricted.data$bioclim19 - restricted.data$LIG.bi19)

historical.variation.data <- historical.variation.data %>% drop_na(bio1.LIG.diff)

#Run a PCA of the difference removing the variation variables (4 and 15)

#For LIG
PCA.LIG.bioclim <- prcomp(historical.variation.data[c(
'bio1.LIG.diff',
'bio2.LIG.diff',
'bio3.LIG.diff',
'bio5.LIG.diff',
'bio6.LIG.diff',
'bio7.LIG.diff',
'bio8.LIG.diff',
'bio9.LIG.diff',
'bio10.LIG.diff',
'bio11.LIG.diff',
'bio12.LIG.diff',
'bio13.LIG.diff',
'bio14.LIG.diff',
'bio16.LIG.diff',
'bio17.LIG.diff',
'bio18.LIG.diff',
'bio19.LIG.diff')],
                      scale = TRUE, center = TRUE)

#Create table with PC1 and PC2
as.data.frame(PCA.LIG.bioclim$rotation[,1:2]) %>% `colnames<-`(c("PC1.LIG", "PC2.LIG")) %>% pander()
```

```{r, fig.width=8, fig.height=5}
#Now we can predict the PCA results: 
PCA.LIG.predictions <- as.data.frame(predict(PCA.LIG.bioclim)*(-1)) #So that higher numbers mean more variation
PCA.LIG.predictions <- rename(PCA.LIG.predictions, PC1.LIG = PC1,
       PC2.LIG = PC2)

#Bind them to dataframe, taking only the first two PCAs
historical.variation.data <- cbind(historical.variation.data, PCA.LIG.predictions[1:2])

#Now to the restricted dataframe
restricted.data <- right_join(restricted.data, historical.variation.data %>% dplyr::select(binomial, PC1.LIG, PC2.LIG), by = 'binomial')

PCA.plot <- historical.variation.data %>% ggplot(aes(x = PC1.LIG, y= PC2.LIG))+
  geom_point(shape = 21)+
  theme_minimal()

PCA.plot.m <- ggExtra::ggMarginal(
  p = PCA.plot,
  type = 'density',
  margins = 'both',
  size = 5,
  colour = 'black',
  fill = 'gray'
)

grid.arrange(PCA.plot.m)
```

**Figure S2:** The distribution of the two PCA compontents for the absolute difference in bioclimatic variables between today and the LIG.

##Correlations between environmental predictors

We can check the correlations between environmental predictors that we will use in our model. Specifically we want to know whether the following environmental variables are correlated:

 + Range size  
 + Short temporal variability of temperature (mean BIOCLIM4)  
 + Spatial variability of temperature (PCA1) [residual.PCA1]  
 + Long-term variability of temperature (LIG)  
 + NPP

To check some of the correlations between the environmental predictors to be used in the PGLS models we can inspect the following correlation plot with the correlation value plotted:

```{r, fig.width=10, fig.height=10}
#Draw plot
restricted.data$log.range.size <- log(restricted.data$range.size.m2)
pairs(restricted.data[,c("log.range.size", "bioclim4", "residuals.PC1", "PC1.LIG", "NPP")], lower.panel=panel.smooth, upper.panel=panel.cor)
```

**Figure S3:** The highest correlation is between Long term temperature variation and seasonal variation. However this correlation is not excessively high (0.53) so we can be satisfied in including these predictors within our model.

##Sexual Dimorphism

The sexual dimorphism dataset looks to have an overdispersed distribution (see phylogenic plot in manuscript). Unfortunately, transformations do not greatly improve the distribution. Our model fit is expected to be reduced by this but further transformations are not obvious and would risk problems of interperetation.

@@

#Analysis

##Simple speciation measures: Diversification rate (DR) and Node Density (ND)

Diversification rates can be simply estimated from thew 'tippiness of the phylogeny'. Here a log equal splits (also called DR) value can be achieved that is weighted to more recent diversification. It was used by @Jetz_2012 and recently (in comparison with BAMM) in a study ivestigation bird diversity alongside elevation [@quintero_2018]. We can look at speciation rate using ES-Sim adapted by @Harvey_essim_2017 and found here https://github.com/mgharvey/ES-sim.

The functions that generate these tip rates estimates are: ``calculate.log.es``, ``calculate.nd`` and ``calculate.tb``. Here they are run on 100 trees plus a full MCC tree of passerine birds based on 2500 samples of the posterior [@Jetz_2012]. This was done using the ``phanghorn`` [@phangorn_2010].

```{r}
#Obtain DR (log(es)) estimates, by calling the calculate.dr function
passerine.trees <- read.nexus('bamm_extinction/data/trees/Passerine_Trees_Full.nex')

#DR, ND and TB can be calculated on the 100 trees by:

# es.list <- sapply(passerine.trees, calculate.log.es)
# nd.list <- sapply(passerine.trees, calculate.nd)
# tb.list <- sapply(passerine.trees, calculate.tb)

#To avoid re-running the time-consuming rates, we can save them and load them: 
es.list <- readRDS("data/es.list.rds")
nd.list <- readRDS("data/nd.list.rds")
tb.list <- readRDS("data/tb.list.rds")

#Also load in our MCC tree
MCC.passerine <- read.tree('data/MCC.passerine.tre') #Tree based on 2500 samples of posterior
```

To gain an estimate of variation between trees we can take the values and obtain a mean, variance and coeeficient of variation (CV). We obtain a CV value to account for unequal variation at higher estimates. The CV is expressed as the ratio of the standard deviation ($\sigma$) to the mean ($\mu$) (converting to a percentage not needed):

$C_V = \frac{\sigma}{\mu} \times 100$

Furthermore, we can account for overdispersion by obtaining a transformed version of CV. In such a case we can calculate CV for a log-normal distribution:

${\widehat {c_{\rm {v}}}}_{\rm {ln}}={\sqrt {\mathrm {e} ^{s_{\rm {ln}}^{2}}-1}}$

Given that our _ES_ value is log-transformed, the calculation of ${\widehat {c_{\rm {v}}}}_{\rm {ln}}$ is perhaps more ideal. As for _ND_ we can see from the a ``qqnorm`` plot that value of ND is normally distributed, therefore the regular CV is more appropriate for this distribution.

```{r, eval = FALSE}
#For ND
#transpose the list so that the elements are the species
t.nd.list <- nd.list %>% purrr::transpose()
#not too sure why this works but turn the nested list into a dataframe
nd.values <- as.data.frame(t(sapply(t.nd.list, function(x) sapply(x, function(x) (x)))))
nd.values <- nd.values %>% tibble::rownames_to_column("binomial")
nd.summary <- melt(nd.values, id.vars = "binomial") %>% group_by(binomial) %>% summarise(
  mean.nd = mean(value), #use normal mean, not a rate
  var.nd = var(value), #Don't really need this if we use CV
  CV.nd = (sd(value)/mean(value)) #Given normal distribution of ND, standard CV formula is appropriate
)

#For ES
#transpose the list so that the elements are the species
t.es.list <- es.list %>% purrr::transpose()
#not too sure why this works but turn the nested list into a dataframe
es.values <- as.data.frame(t(sapply(t.es.list, function(x) sapply(x, function(x) (x)))))
es.values <- es.values %>% tibble::rownames_to_column("binomial")
es.summary <- melt(es.values, id.vars = "binomial") %>% group_by(binomial) %>% summarise(
  mean.loges = log(1/mean(1/exp(value))), #use harmonic mean for rates and log-transform post calculation of HM
  var.loges = var(value), #Don't really need this if we use CV
  CV.loges = sqrt(exp(var(value)) - 1) #CV for log-normal distribution of ES
)

#For TB
#transpose the list so that the elements are the species
t.tb.list <- tb.list %>% purrr::transpose()
#not too sure why this works but turn the nested list into a dataframe
tb.values <- as.data.frame(t(sapply(t.tb.list, function(x) sapply(x, function(x) (x)))))
tb.values <- tb.values %>% tibble::rownames_to_column("binomial")
tb.summary <- melt(tb.values, id.vars = "binomial") %>% group_by(binomial) %>% summarise(
  mean.tb = mean(value), #use normal mean; not a rate
  var.tb = var(value), #Don't really need this if we use CV
  CV.tb = (sd(value)/mean(value)) #Given normal distribution of tb, standard CV formula is appropriate
)

MCC.nd.df <- as.data.frame(calculate.nd(MCC.passerine)) %>% tibble::rownames_to_column("binomial")
MCC.dr.df <- as.data.frame(calculate.log.es(MCC.passerine)) %>% tibble::rownames_to_column("binomial")

#Join the dataframes
dr.summary <- full_join(nd.summary, es.summary, by = "binomial")
dr.summary <- full_join(dr.summary, tb.summary, by = "binomial")
dr.summary <- full_join(dr.summary, MCC.nd.df, by = "binomial")
dr.summary <- full_join(dr.summary, MCC.dr.df, by = "binomial")

dr.summary <- dr.summary %>% rename(TipLabel = binomial, MCC.ND = nd, MCC.DR = es)

saveRDS(dr.summary, 'data/dr.summary.rds')
```

```{r}
dr.summary <- readRDS('data/dr.summary.rds')
#Plot a summary of logES and TB with the DRs with their weightings
dr.plot <- dr.summary %>% ggplot(aes(x = mean.loges, y = mean.nd, 
                          fill = (1/(CV.nd))/sum((1/(CV.nd)))*100, 
                      size = (1/(CV.loges))/sum((1/(CV.loges)))*100))+
  geom_point(shape = 21, alpha = 0.5)+
  theme_minimal()+
  geom_errorbarh(aes(xmin = mean.loges - 0.674*sqrt(var.loges), xmax = mean.loges + 0.674*sqrt(var.loges)), 
                 size = 0.015)+
  geom_errorbar(aes(ymin = mean.nd - 0.674*sqrt(var.nd), ymax = mean.nd + 0.674*sqrt(var.nd)), 
                size = 0.015)+
  theme(legend.position="bottom")+
  labs(size = 'Weight (%) [using DR]', y='Node Density [ND]', x= 'Log Diversification Rate [DR]', fill = 'Weight (%) [using ND] \n')+
  scale_fill_distiller(guide = "colorbar", palette = "RdPu", direction = 1, breaks=c(0,0.015,.03),
                           limits=c(0,0.03))
marginal.dr <-ggMarginal(
  p = dr.plot,
  type = 'density',
  margins = 'both',
  size = 5,
  colour = '#0000009C',
  fill = '#D12E6769'
)
```

```{r, fig.width=8, fig.height=6}
grid.newpage()
grid.draw(marginal.dr)
```
**Figure S4:** $\lambda_{DR}$ and $\lambda_{ND}$ are correlated in their estimation of species rate. Furthermore, we see similar variation across each species in the estimates across 100 trees as the inverse of the coefficeint of variation (weight) is approximately equivalent across all values of ND and DR (colours and size). Notably, these estimates speciation rate are relatively uncertain across all species, with 50 % CIs in both axes plotted.

Combine the estimates of DR and ND with the restricted dataframe containing all the other variables. We can inspect the trend in SDi in relation to the speciation rates.

```{r, warning=FALSE, fig.width=8, fig.height=10}
restricted.data <- left_join(restricted.data, dr.summary, by = 'TipLabel')

es.plot <- restricted.data %>% ggplot(aes(y = mean.loges, x = SDi, fill = mean.loges))+
geom_point(shape = 21, size = 1.5)+
geom_smooth(show.legend = FALSE, color = "grey20", method = "lm")+
scale_fill_distiller(palette = "YlOrBr", direction = 1, guide = FALSE)+
ylab("mean.DR")+
theme_minimal()

nd.plot <- restricted.data %>% ggplot(aes(y = mean.nd, x = SDi, fill = mean.nd))+
geom_point(shape = 21, size = 1.5)+
geom_smooth(show.legend = FALSE, color = "grey20", method = "lm")+
scale_fill_distiller(palette = "Greens", direction = 1, guide = FALSE)+
theme_minimal()

tb.plot <- restricted.data %>% ggplot(aes(y = mean.tb, x = SDi, fill = mean.tb))+
geom_point(shape = 21, size = 1.5)+
geom_smooth(show.legend = FALSE, color = "grey20", method = "lm")+
scale_fill_distiller(palette = "PuBu", direction = 1, guide = FALSE)+
theme_minimal()
